Anomaly detection with kernel density estimation on manifolds

### 2022-03-22 paper/paper.tex
Riemannian metric, geodesics, exp and log maps

A geodesic is a smooth
curve that locally joins their points along the shortest path. All
the geodesics passing through p are called radial geodesics. The
local coordinates defined by the chart ðU ; exp1 Þ are called p
(Riemannian) normal coordinates with center p.~Note that normal coordinates preserve the distances on radial geode- sics. For example, unfolding a sphere onto a plane in normal coordinates can preserve the distances on great circles, like paring an orange. Note that for a manifold with complex topology, for example, a torus, multiple normal coordinate charts, rather than one single chart, are needed to cover the whole manifold.

@@

@@

For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\). \(\gamma_1\) and \(\gamma_2\) are equivalent if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
\[
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
\]
A \emph{tangent vector} to \(M\) at \(p\) is any equivalent class of the differentiable curves initialized at \(p\).
The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\). The tangent space is vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the charts \(\varphi\) locally at \(p\). The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
Tangent vectors can also be seen as the directionnal derivatives at \(p\). For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a basis of the tangent space \(T_pM\).

The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the cotangent space with corresponding cotangent vector \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).

Define the directional derivative \(D_v\) in the direction of \(v\) as
\[
\forall f \in C^\infty(M): D_v(f)
\]

\hypertarget{riemannian-metrics}{%
\subsection{Riemannian metrics}\label{riemannian-metrics}}

Let \((\mathcal{M}, g)\) be a \(d\)-dimensional Riemannian manifold without boundary. Denote \(d_g\) the Riemannian distance. Assume that \((\mathcal{M}, g)\) is complete, i.e.~\((\mathcal{M}, d_g)\) is a complete metric space, and that the injectivity radius of \((\mathcal{M}, g)\), \(inj_g\mathcal{M}\), is strictly positive.

Denote \(B_s(p)\) the normal ball in \((\mathcal{M}, g)\) centered at point \(p\) with raidus \(s\).

\(d_g\) is induced by \(g\) and by the injectivity radius of \((\mathcal{M}, g)\).

normal ball

@@

embedding pushforward metric \autocite{Perrault-Joncas2013-pq}

@@

For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional.

\hypertarget{isometry}{%
\subsection{Isometry}\label{isometry}}

>>>>

@@

Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism. \(\varphi(N)\) is called an embedded submanifold of \(M\) with the subspace topology.

@@

For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\). \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
\[
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
\]
A \emph{tangent vector} to \(M\) at \(p\) is any equivalent class of the differentiable curves initialized at \(p\).
The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\). The tangent space is vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the charts \(\varphi\) locally at \(p\). The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
Tangent vectors can also be seen as the directional derivatives at \(p\). For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space \(T_pM\).

The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).

Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\). The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
\[
<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
\]
where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).

\hypertarget{riemannian-metric-and-geodesic-distance}{%
\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}

A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\) symmetric positive definite and varies smoothly at \(p\). Generally, we omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
The inner product for two vectors \(u, v \in T_pM\) is written as \(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).

The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
\(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
For a curve \(\gamma: I \rightarrow M\), the length of the curve is
\[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
where \(\gamma(I) \subset U\). The volume \(W \subset U\) is defined as
\[
Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
\]
which is also called the \emph{Riemannian measure} on \(M\).

The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold. A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\),
the second-order ordinary differential equation is satisfied,
\[
\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
\]
where \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
\[
\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
\]
The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt\) when the curves are defined over the interval \([0,1]\). The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the \emph{injectivity radius} at \(p \in M\) is the distance to the cut locus. The injectivity radius of \((M,g)\), \(\textit{inj}_gM\) is the infimum of the injectivity radii over all points on the manifold.

@@

Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\). Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\).
Define \(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a differentiable, bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\).
The exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint after covering the length of \(\|u\|\) along the geodesic in one time unit.
The inverse of the exponential map is called the \emph{logarithm map}, \(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to \(q\).
Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).

@@

For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
\[
d\exp_p^*vol(u) = \theta_p(u)du.
\]
